Columnar VCF using Zarr

### 2024-03-28 paper.tex
Cleanup

\subsection{Overview of narrative for draft}

\paragraph{Storing genetic variation data}
High level discussion of
strategies used to store genotype matrix and accompanying metadata.
Emphasis on access by variant, not sample. Introduce Zarr, and
refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
and Fig~\ref{fig-data-storage}B to show that is competitive
with highly specialised approaches.
Note that the FC simulated dataset is probably a best-case
scenario for methods specialised to patterns in the genotype matrix.
Discuss storing additional data, and advantages of the
columnar approach (refs to genomics tools that do the same).

\paragraph{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.
We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations
of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks. We discuss parallelism, and why
bcftools and genozip can make limited use of available cores.
Also that single-machine parallelism is inevitably constrained
by IO (which is why savvy's thread utilisation drops off).
Figure: \ref{fig-whole-matrix-compute}. 
% Also
% supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.

\paragraph{Subsetting the genotype matrix}
Existing work overwhelmingly emphasises accessibility by variant.
We demonstrate that sgkit/zarr is good for getting subsets
in terms of both variants and samples via afdist on various subsets.
Discuss why storing AF values for crude population groupings is
intrinsically bad, and that it should be easy to do calculations
on aribitrary subsets. Also participants regularly drop-out of
large studies like UKB - do you update the AF values each time?

@@

\subsection{Benchmarking}
Full details of the benchmarking done in
Figure~\ref{fig-data-storage} and
Figure~\ref{fig-whole-matrix-compute}.
Quick notes:

\begin{itemize}
\item Dataset is based on subsets of French-Canadian sims. We simplify
for increasingly large subsets, keeping only variant sites. We then
convert to VCF and encode to BCF using
\texttt{tskit vcf | bcftools view -O b}.
\item We used a chunk size of XX for sgkit, after some experimentation.
This gave n files of around x MB each for the ``call\_genotypes`` array.
We used the defaults for savvy and genozip.
\item For the afdist CPU time we measure the sum of the total user and
system times required to execute the full command, as reported by GNU
time. Each tool was instructed to use one thread, where the options
were provided. For genozip, the time required to compute
\texttt{genocat file.genozip | bcftools +af-dist}. Because commands
in the pipeline execute concurrently on mutiple cores, the total CPU time is
greater than the elapsed time.
\item For the threading speedup plot, we report the time required to compute
afdist with 8 threads relative to 1 thread (a speedup of 8 is therefore
the maximum). For savvy, our C++ implementation of afdist calculates
on 8 chunks of genome, and aggegates once they are completed.
Bcftools has a ``--threads`` option, but it only applies to decoding.
\item Where possible in pipelines we use uncompressed BCF
 output \texttt{-Ou} to make processing more efficient (skip printf
and parsing costs, which are substantial).
\item We do not use BCF output in genozip because it doesn't support
it directly, only VCF (supports BCF by pipeing through bcftools).
\end{itemize}

@@

% \begin{figure*}[ht]
% 	\begin{center}
% 		\includegraphics[width=\textwidth]{figures/whole-matrix-compute-supplemental}
% 	\end{center}
% 	\caption{\label{fig-whole-matrix-compute-supplemental}
%     Limitations of vertical scaling on whole-matrix afdist. The elapsed
% wall-clock time required to compute afdist on the whole genotype matrix using
% different numbers of threads and storage backends (top row), and the relative
% speedup obtained from threads as a fraction of optimum speedup (bottom row).
% [TODO more detail and link to fig 1. Also add the times for bcftools and
% genozip in col 1?]
% }
% \end{figure*}

>>>>

@@

\subsection{Benchmarks}
Full details of the benchmarking done in
Figure~\ref{fig-data-storage} and
Figure~\ref{fig-whole-matrix-compute}.
Quick notes:

\begin{itemize}
\item Dataset is based on subsets of French-Canadian sims. We simplify
for increasingly large subsets, keeping only variant sites. We then
convert to VCF and encode to BCF using
\texttt{tskit vcf | bcftools view -O b}.
\item We used a chunk size of XX for sgkit, after some experimentation.
This gave n files of around x MB each for the ``call\_genotypes`` array.
We used the defaults for savvy and genozip.
\item For the afdist CPU time we measure the sum of the total user and
system times required to execute the full command, as reported by GNU
time. Each tool was instructed to use one thread, where the options
were provided. For genozip, the time required to compute
\texttt{genocat file.genozip | bcftools +af-dist}. Because commands
in the pipeline execute concurrently on mutiple cores, the total CPU time is
greater than the elapsed time.
\item Where possible in pipelines we use uncompressed BCF
 output \texttt{-Ou} to make processing more efficient (skip printf
and parsing costs, which are substantial). Following advice
in~\citep{danecek2021twelve}.
\item We do not use BCF output in genozip because it doesn't support
it directly, only VCF (supports BCF by pipeing through bcftools).
\end{itemize}

@@
